Special Constructions in Categories

### 2017-08-24 papers/talk-6/paper-6.tex
writeup6: Erste Korrektur-Iteration

% { _ }
\newcommand{\setCon}[1]{\bigl \{ #1 \bigr \}}
% { _ | _ }
\newcommand{\setMap}[2]{\setCon{#1 \,\big|\, #2}}
% ( _ , _ )
\newcommand{\pair}[2]{\left( #1 , #2 \right)}
% [ _ ]
\newcommand{\class}[1]{\left[ #1 \right]}
% < _ >
\newcommand{\choice}[1]{\left< #1 \right>}

@@

% Title page
\title{Special Constructions in Categories}
\author{Maximilian Wuttke}

@@

\begin{example}[Cartesian Product]
  \label{ex:prod:cart}
  Let $X$ and $Y$ be sets.
  Then we define the \demph{Cartesian Product}
  $$X \times Y := \setMap{\pair{x}{y}}{x \in X, \; y \in Y}. $$
  Also we define the \demph{projection functions} $p_1 \from X \times Y \to X$, $p_2 \from X \times Y \to Y$ in a canonical way:
  \begin{alignat*}{2}
    p_1(x,y) &:= x \\
    p_2(x,y) &:= y
  \end{alignat*}

  Let $A$ be another set and $f_1 \from A \to X, f_2 \from A \to Y$ functions.
  If we want to define a function $\fbar \from A \to X \times Y$, such that
  $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ for every $a \in A$,
  there is only one possible way to define $\fbar \from A \to X \times Y$:
  $$\fbar(a) := \pair{f_1(a)}{f_2(a)}.$$
  We have $p_1(\fbar(a)) = p_1(f_1(a), f_2(b)) = f_1(a)$ and $p_2(\fbar(a)) = p_2(f_1(a), f_2(a)) = f_2(a)$.
  If we have another $\ftilde \from A \to X \times Y$ with $p_1(\ftilde(a)) = f_1(a)$ and $p_2(\ftilde(a)) = f_2(a)$ we show $\fbar(a) = \ftilde(a)$ for every $a \in A$:
  $$\ftilde(a) = \pair{f_1(a)}{f_2(a)} = \pair{p_1(\ftilde(a)}{p_2(\ftilde(a))} = \ftilde(a).$$
  The last step is because we have $\pair{p_1(z)}{p_2(z)} = x$ for every $z \in X \times Y$.
\end{example}

@@

\begin{definition}[Product]
  \label{def:prod}
  Let $\cat{A}$ be a category and $X, Y \in \cat{A}$ objects.
  A \demph{product} of $X$ and $Y$ consists of an object $P$ and maps
  $ \xymatrix{
    X &P \ar[l]_{p_1} \ar[r]^{p_2} & Y
  } $
  with the property that for all objects and maps
  $ \xymatrix{
    X &A \ar[l]_{f_1} \ar[r]^{f_2} & Y
  } $
  in $\cat{A}$, there exists a unique map $\bar{f}\from A \to P$ such that
  \[ \xymatrix{
    & A \ar[ldd]_{f_1} \ar@{.>}[d]|{\bar{f}} \ar[rdd]^{f_2} & \\
    & P \ar[ld]^{p_1} \ar[rd]_{p_2} & \\
    X & & Y
  } \]
  commutes. The maps $p_1$ and $p_2$ are called the \demph{projections}.
\end{definition}

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be a product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.

  Lets reconsider the definition of the \emph{product}:
  For every object $A \in \cat{A}$ and arrows $f_1 \from A \to X, f_2 \from A \to Y$ we get that there exists a unique map
  $\bar{f} \from A \to P$, such that
  \[ \xymatrix{
      &A \ar[ldd]_{f_1} \ar@{.>}[d]|{\bar{f}\vphantom{\bar{\bar{f}}}}
      \ar[rdd]^{f_2}&       \\
              &P \ar[ld]^{p_1} \ar[rd]_{p_2}                  &       \\
      X       &                                               &Y
  } \]
  commutes.
  Now we can instantiate $A$ with $P'$, $f_1$ with $p_1$, $f_2$ with $p_2$. By analogous instantiations we get the commuting diagram
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\bar{f}} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\tilde{f}} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \tilde{f}, \bar{f}$. With the inner triangles we get:
  \begin{alignat*}{2}
    p_1  \of \tilde{f} &= p'_1 &\qquad p_2  \of \tilde{f} &= p'_2 \\
    p'_1 \of \bar{f}   &= p_1  &       p'_2 \of \bar{f}   &= p_2
  \end{alignat*}
  By substitution we get
  \begin{alignat*}{2}
    p_1   \of (\tilde{f} \of \bar{f}) &= p_1  &\qquad p_2  \of (\tilde{f} \of \bar{f}) &= p_2 \\
    p'_1  \of (\bar{f} \of \tilde{f}) &= p'_1 &       p'_2 \of (\bar{f} \of \tilde{f}) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \tilde{f} \of \bar{f}$, thus $\tilde{f} \of \bar{f} = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \bar{f} \of \tilde{f}$, thus $\bar{f} \of \tilde{f} = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

\begin{example}[Reversed Cartesian product]
  \label{ex:prod:cart2}
  For sets we can define another product, where we reverse the order of the elements, i.~e.:
  \begin{alignat*}{2}
    X \times' Y &:= \setMap{\pair{y}{x}}{x \in X, \; y \in Y} \\
    p'_1(x,y)   &:= y \\
    p'_2(x,y)   &:= x
  \end{alignat*}
  Now $X \times Y$ and $X \times' Y$ are isomorphic with the involution $i(a, b) := \pair{b}{a}$.
\end{example}

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field with the operations $+_K$ and $\cdot_K$ and $V, W$ vector spaces over this field with the operations $+_V, \cdot_V, +_W, \cdot_W$.
  Then we define\footnote{One should verify that the defined structure is indeed a vector field.}
  the \demph{direct sum} $V \oplus W$ of $V$ with $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ and with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 +_V v_2}{w_1 +_W w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot_V v}{\alpha \cdot_W w} &     & \forall v \in V, \forall w \in W
  \end{alignat*}
  We can now show that $U \oplus V$ is the product of $U$ and $V$ in $\Vect$.
  Therefore we can forget that $U$ and $V$ are vector spaces and follow the proof in example \ref{ex:prod:cart}.
\end{example}

@@

We can again use this knowledge to define coproducts in arbitrary categories.

\begin{definition}[Coproduct]
  \label{def:coproduct}
  Let $\cat{A}$ be a category and $X, Y \in \cat{A}$.
  A \demph{coproduct} of $X$ and $Y$ consists of an object $S$ and maps
  $ \xymatrix{
    X \ar[r]^{p_1} & S & Y \ar[l]_{p_2}
  } $
  with the property that for all objects and maps
  $ \xymatrix{
    X \ar[r]^{f_1} & A & Y \ar[l]_{f_2}
  } $
  in $\cat{A}$, there exists a unique map $\bar{f}\from S \to A$ such that
  \[ \xymatrix{
    X \ar[rdd]_{f_1}\ar[rd]^{p_1} & & Y \ar[ldd]^{f_2} \ar[ld]_{p_2} \\
    & S \ar@{.>}[d]|{\bar{f}} & \\
    & A &
  } \]
  commutes. The maps $p_1$ and $p_2$ are called the \demph{coprojections}.
\end{definition}

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In the example \ref{ex:prod:vspace} we have seen that the direct sum of two vectorspaces $U$ and $V$ is the product of $V$ and $W$ in $\Vect$.
  We will now see that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in the examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are vector morphims we have:
  \begin{alignat*}{4}
    \fbar(p_1(u)) &= \fbar(u, 0) &= f_1(u) + f_2(0) &= f_1(u) \\
    \fbar(p_2(v)) &= \fbar(0, v) &= f_1(0) + f_2(v) &= f_2(v).
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphims with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphims with $\ftilde(p_1(u)) = f_1(u)$ and $\ftilde(p_2(v)) = f_2(v)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$ \fbar(v,w) = f_1(v) + f_2(w)\ = \ftilde(p_1(v)) + \ftilde(p_2(w)) = \ftilde(p_1(v) + p_2(w)) = \ftilde(v, w). $$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the the category $\Vect$.
\end{example}

@@

%% XXX Weiterhin erst Beispiel und dann Definition oder erst Definition und dann Beispiel?
Now We present abstract concepts and give examples.

\begin{definition}[Equalizer]
  \label{def:equa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  An \demph{equalizer} of this diagram is a commuting diagram called \demph{fork} \\
  $ \xymatrix{
    E \ar[r]^i & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  with the property that for all other forks
  $ \xymatrix{
    A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  in $\cat{A}$, there exists a unique map $\bar{f}\from A \to E$ such that
  \[ \xymatrix{
    A \ar[rd]^f \ar@{.>}[dd]|{\bar{f}} & & \\
    & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \\
    E \ar[ru]_i & &
  } \]
  commutes.
\end{definition}

@@

\begin{definition}[Coequalizer]
  \label{def:coequa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  A \demph{coequalizer} of this diagram is a commuting diagram called \demph{cofork} \\
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  with the property that for all other coforks
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  in $\cat{A}$, there exists a unique map $\bar{f}\from C \to A$ such that
  \[ \xymatrix{
    & & A  \\
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
    & & C \ar@{.>}[uu]|{\bar{f}}
  } \]
  commutes.
\end{definition}

For our next example we need the definition of the reflexive transitive closure.

\begin{definition}[Symmetric transitive closure]
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {(\hat{R})^n}$
  where ${\hat R}^n$ is defined recursively with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := R_n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in Y} \subseteq Y^2$ the relation of values in $Y$ that correspond from the same $x \in X$.
  Let $\sim \subseteq Y^2$ be the symmetric transitive closure of $R$.
  Now we define $C := Y / \sim = \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ ``does not care'' about equal values, i.~e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot$ that returns for every equivalence class in $\sim$ a member of it, i.~e.
  $\forall c \in C, \forall y \in e, \choice{e} \sim y$.
  Therefore we also know $\forall c \in C, \class {\choice c} = c$.

  Now we can define $\fbar \from C \to A$ by $\fbar(e) := f(\choice e)$.
  We now have  $\fbar(i(y)) = f(\choice{\class y}) = f(y)$ because $\choice{\class y} \sim y$ (by reflexivity of $\sim$).
  Now we only have to show that this $\fbar$ is unique. So we assume another function $\ftilde \from C \to A$ with $\ftilde \of i = f$. But then we have
  $$ \fbar(c) = f(\choice c) = \ftilde(i(\choice c)) = \ftilde(\class{\choice c}) = \ftilde(c).$$
\end{example}

@@

\begin{definition}[Pullback]
  \label{def:pullback}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pullback} of this diagram is a \demph{pullback square}
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  with the property that for all other pull back squares
  \[ \xymatrix{
    A \ar[r]_{f_2} \ar[d]^{f_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  in $\cat{A}$, there exists a unique map $\bar{f}\from A \to P$ such that
  \[ \xymatrix{
    A \ar[drr]^{f_2} \ar[ddr]_{f_1} \ar@{.>}[dr]|{\bar{f}} & & \\
    & P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    & X \ar[r]^s & Z
  } \]
  commutes.
\end{definition}

>>>>

% { _ }
\newcommand{\setCon}[1]{\bigl \{ #1 \bigr \}}
% { _ | _ }
\newcommand{\setMap}[2]{\setCon{#1 \,\big|\, #2}}
% ( _ , _ )
\newcommand{\pair}[2]{\left( #1 , #2 \right)}
% [ _ ]
\newcommand{\class}[1]{\left[ #1 \right]}
% < _ >
\newcommand{\choice}[1]{\left< #1 \right>}
% explanation over relation sign
\newcommand{\explainRel}[2]{\stackrel{\text{#1}}{#2}}

@@

% Title page
\title{Special Constructions in Categories}
\subtitle{Examples for limits and colimits}

\author{Maximilian Wuttke}

@@

\begin{example}[Cartesian Product]
  \label{ex:prod:cart}
  Let $X$ and $Y$ be sets.
  Then we define the \demph{Cartesian Product}
  $$X \times Y := \setMap{\pair{x}{y}}{x \in X, \; y \in Y}.$$
  Also we define the \demph{projection functions} $p_1 \from X \times Y \to X$, $p_2 \from X \times Y \to Y$ in a canonical way:
  \begin{alignat*}{2}
    p_1(x,y) &:= x \\
    p_2(x,y) &:= y
  \end{alignat*}

  Let $A$ be another set and $f_1 \from A \to X, f_2 \from A \to Y$ functions.
  If we want to define a function $\fbar \from A \to X \times Y$, such that
  $p_1(\fbar(a)) = f_1(a)$ and $p_2(\fbar(a)) = f_2(a)$ for every $a \in A$ holds,
  there is only one possible way to define $\fbar \from A \to X \times Y$:
  $$\fbar(a) := \pair{f_1(a)}{f_2(a)}.$$
  By construction we have $p_1(\fbar(a)) = p_1(f_1(a), f_2(b)) = f_1(a)$ and $p_2(\fbar(a)) = p_2(f_1(a), f_2(a)) = f_2(a)$.
  If we have another $\ftilde \from A \to X \times Y$ with $p_1(\ftilde(a)) = f_1(a)$ and $p_2(\ftilde(a)) = f_2(a)$ we show $\fbar(a) = \ftilde(a)$ for every $a \in A$:
  $$\fbar(a) = \pair{f_1(a)}{f_2(a)} = \pair{p_1(\ftilde(a)}{p_2(\ftilde(a))} = \ftilde(a).$$
  The last step is because we have $\pair{p_1(z)}{p_2(z)} = z$ for every $z \in X \times Y$.
\end{example}

@@

\begin{definition}[Product]
  \label{def:prod}
  Let $\cat{A}$ be a category and $X, Y \in \cat{A}$ objects.
  A \demph{product} of $X$ and $Y$ consists of an object $P$ and maps
  $ \xymatrix{
    X &P \ar[l]_{p_1} \ar[r]^{p_2} & Y
  } $
  with the property that for all objects and maps
  $ \xymatrix{
    X &A \ar[l]_{f_1} \ar[r]^{f_2} & Y
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from A \to P$ such that
  \[ \xymatrix{
    & A \ar[ldd]_{f_1} \ar@{.>}[d]|{\fbar} \ar[rdd]^{f_2} & \\
    & P \ar[ld]^{p_1} \ar[rd]_{p_2} & \\
    X & & Y
  } \]
  commutes. The maps $p_1$ and $p_2$ are called the \demph{projections}.
\end{definition}

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be a product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.
  We apply definition \ref{def:prod} multiple times and get the following commuting diagram:
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\fbar} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\ftilde} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \fbar, \ftilde$. With the inner triangles we get:
  \begin{alignat*}{2}
    p_1  \of \ftilde &= p'_1 &\qquad p_2  \of \ftilde &= p'_2 \\
    p'_1 \of \fbar   &= p_1  &       p'_2 \of \fbar   &= p_2
  \end{alignat*}
  By substitution we get:
  \begin{alignat*}{2}
    p_1   \of (\ftilde \of \fbar) &= p_1  &\qquad p_2  \of (\ftilde \of \fbar) &= p_2 \\
    p'_1  \of (\fbar \of \ftilde) &= p'_1 &       p'_2 \of (\fbar \of \ftilde) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \ftilde \of \fbar$, thus $\ftilde \of \fbar = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \fbar \of \ftilde$, thus $\fbar \of \ftilde = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

\begin{example}[Reversed Cartesian product]
  \label{ex:prod:cart2}
  For sets we can define another product, where we reverse the order of the elements, i.~e.:
  \begin{alignat*}{2}
    X \times' Y &:= \setMap{\pair{y}{x}}{x \in X, \; y \in Y} \\
    p'_1(y,x)   &:= x \\
    p'_2(y,x)   &:= y
  \end{alignat*}
  Now $X \times Y$ and $X \times' Y$ are isomorphic with the involution $i(a, b) := \pair{b}{a}$.
\end{example}

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field with the operations $+_K$ and $\cdot_K$ and $V, W$ vector spaces over this field with the operations $+_V, \cdot_V, +_W, \cdot_W$.
  Then we define\footnote{One should verify that the defined structure is indeed a vector field.}
  the \demph{direct sum} $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ as a new vector field with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 +_V v_2}{w_1 +_W w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot_V v}{\alpha \cdot_W w} &     & \forall v \in V, \forall w \in W
  \end{alignat*}
  We can show that $U \oplus V$ is the product of $U$ and $V$ in $\Vect$.
  The proof follows the same structure as in example \ref{ex:prod:cart}.
\end{example}

@@

Again we can use this knowledge to define coproducts in arbitrary categories.

\begin{definition}[Coproduct]
  \label{def:coproduct}
  Let $\cat{A}$ be a category and $X, Y \in \cat{A}$.
  A \demph{coproduct} of $X$ and $Y$ consists of an object $S$ and maps
  $ \xymatrix{
    X \ar[r]^{p_1} & S & Y \ar[l]_{p_2}
  } $
  with the property that for all objects and maps
  $ \xymatrix{
    X \ar[r]^{f_1} & A & Y \ar[l]_{f_2}
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from S \to A$ such that
  \[ \xymatrix{
    X \ar[rdd]_{f_1}\ar[rd]^{p_1} & & Y \ar[ldd]^{f_2} \ar[ld]_{p_2} \\
    & S \ar@{.>}[d]|{\fbar} & \\
    & A &
  } \]
  commutes. The maps $p_1$ and $p_2$ are called the \demph{coprojections}.
\end{definition}

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In the example \ref{ex:prod:vspace} we have seen that the direct sum of two vector spaces $U$ and $V$ is the product of $V$ and $W$ in $\Vect$.
  We will now see that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in the examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are vector morphism we have:
  \begin{alignat*}{4}
    \fbar(p_1(u)) &= \fbar(u, 0) &= f_1(u) + f_2(0) &= f_1(u) \\
    \fbar(p_2(v)) &= \fbar(0, v) &= f_1(0) + f_2(v) &= f_2(v).
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphism with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphism with $\ftilde(p_1(u)) = f_1(u)$ and $\ftilde(p_2(v)) = f_2(v)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$ \fbar(v,w) = f_1(v) + f_2(w)\ = \ftilde(p_1(v)) + \ftilde(p_2(w)) = \ftilde(p_1(v) + p_2(w)) = \ftilde(v, w). $$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the category $\Vect$.
\end{example}

@@

%% XXX Weiterhin erst Beispiel und dann Definition oder erst Definition und dann Beispiel?
Now we present more abstract concepts and give examples.

\begin{definition}[Equalizer]
  \label{def:equa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  An \demph{equalizer} of this diagram is a commuting diagram called \demph{fork}
  $ \xymatrix{
    E \ar[r]^i & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  with the property that for all other forks
  $ \xymatrix{
    A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from A \to E$ such that
  \[ \xymatrix{
    A \ar[rd]^f \ar@{.>}[dd]|{\fbar} & & \\
    & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \\
    E \ar[ru]_i & &
  } \]
  commutes.
\end{definition}

@@

\begin{definition}[Coequalizer]
  \label{def:coequa}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  } $
  a diagram in $\cat{A}$.
  A \demph{coequalizer} of this diagram is a commuting diagram called \demph{cofork} \\
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  with the property that for all other coforks
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  in $\cat{A}$, there exists a unique map $\fbar\from C \to A$ such that
  \[ \xymatrix{
    & & A  \\
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[ru]^f \ar[rd]_i \\
    & & C \ar@{.>}[uu]|{\fbar}
  } \]
  commutes.
\end{definition}

For our next example we need the definition of the symmetric reflexive transitive closure.

\begin{definition}[Symmetric reflexive transitive closure]
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the symmetric closure of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is defined recursively with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := R_n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the \emph{least equivalence relation} containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in Y} \subseteq Y^2$.
  Let $\sim \; \subseteq Y^2$ be the symmetric reflexive transitive closure of $R$.
  Now we define $C := Y / \sim = \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ ``does not care'' about equal values, i.~e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot \from C \to Y$ that returns for every equivalence class in $\sim$ a member of it, i.~e.
  $\forall c \in C, \forall y \in Y, y \in c \Leftrightarrow y \sim \choice{c}$
  which is equivalent to $\forall c \in C, \class {\choice c} = c$.

  Now we can define $\fbar \from C \to A$ by $\fbar(e) := f(\choice e)$.
  We now have  $\fbar(i(y)) = f(\choice{\class y}) = f(y)$ because $y \sim \choice{\class y}$ (by the choice axiom and reflexivity of $\sim$).
  Now we only have to show that this $\fbar$ is unique in respect to that property.
  So we assume another function $\ftilde \from C \to A$ with $\ftilde \of i = f$. But then we have
  $$ \fbar(c) \explainRel{Def}= f(\choice c) \explainRel{Ass.}= \ftilde(i(\choice c)) \explainRel{Def}= \ftilde(\class{\choice c}) \explainRel{Choice}= \ftilde(c).$$
\end{example}

@@

\begin{definition}[Pullback]
  \label{def:pullback}
  Let $\cat{A}$ be a category and
  $ \xymatrix{
    X \ar[r]^s & Z & Y\ar[l]_t \\
  } $
  a diagram in $\cat{A}$.
  A \demph{pullback} of this diagram is a \demph{pullback square}
  \[ \xymatrix{
    P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  with the property that for all other pull back squares
  \[ \xymatrix{
    A \ar[r]_{f_2} \ar[d]^{f_1} & Y \ar[d]_{t} \\
    X \ar[r]^s & Z
  } \]
  in $\cat{A}$, there exists a unique map $\fbar\from A \to P$ such that
  \[ \xymatrix{
    A \ar[drr]^{f_2} \ar[ddr]_{f_1} \ar@{.>}[dr]|{\fbar} & & \\
    & P \ar[r]_{p_2} \ar[d]^{p_1} & Y \ar[d]_{t} \\
    & X \ar[r]^s & Z
  } \]
  commutes.
\end{definition}
